---
title: "Predicting Wine Quality"
author: "Ben Jepson"
format: 
  html:
    toc: true
    code-fold: true
engine: knitr
---

```{r}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message=FALSE, dpi = 200, fig.width=10, fig.height=7)
```

```{r}
library(tidyverse)
library(tidymodels)
library(gt)
```

```{r}
#  red wine
red <- read_delim("winequality-red.csv", delim = ";") %>%
  mutate(type = "red") #red wine

#  white wine
white <- read_delim("winequality-white.csv", delim = ";") %>%
  mutate(type = "white") #white wine

wine <- bind_rows(red, white) # easy combine datasets because same structure

rm(red)
rm(white)
```



# Introduction - Wine Quality Prediction

For this project I'm using the wine quality data from the UCI Machine Learning Repository (https://archive.ics.uci.edu/dataset/186/wine+quality)

This dataset has been used in many projects before - there are over 6000 samples of wine with 11 chemicals measured, and quality rated 0-10. I am using it because it is similar in many ways to work I have done in the food industry, combining chemical and rating data to understand a product. 

## Objectives

  - Explore the dataset
  
  - Compare red vs. white wines
  
  - Build and evaluate regression and machine learning models to predict quality
  
  - Show my thought process throughout
  
# Exploratory Data analysis

The dataset provided from the UCI repository does not have missing values. I will explore the data for insights to get started.

### Summary Statistics

```{r}
wine %>% 
  group_by(type) %>% 
  summarize(
    n = n(),
    mean_quality = round(mean(quality),2),
    sd_quality = round(sd(quality),2),
    min_quality = min(quality),
    max_quality = max(quality)
  ) %>% 
    knitr::kable(caption = "wine quality summary")
```

I notice there are about 3 times as many white as red samples. Average quality looks similar, but the standard deviation of quality is large relative to the scale (will end up covering much of the range used). The wines only use 3-9 of the 0-10 scale.

```{r}

wine %>%
  pivot_longer(cols = -c(type, quality), names_to = "feature", values_to = "value") %>%
  group_by(type, feature) %>%
  summarize(
    mean = mean(value),
    sd   = sd(value),
    min  = min(value),
    max  = max(value),
    .groups = "drop"
  ) %>%
  mutate(across(c(mean, sd, min, max), ~round(.x, 2))) %>%
  pivot_wider(
    names_from = type,
    values_from = c(mean, sd, min, max),
    names_glue = "{type}_{.value}"
  ) %>%
  arrange(feature) %>%
  gt(rowname_col = "feature") %>%
  tab_header(
    title = "Red vs. White Wine Summary",
    subtitle = "Mean, SD, Min, and Max for Each Predictor"
  ) %>%
  cols_label(
    red_mean = "Red Mean", white_mean = "White Mean",
    red_sd   = "Red SD",   white_sd   = "White SD",
    red_min  = "Red Min",  white_min  = "White Min",
    red_max  = "Red Max",  white_max  = "White Max"
  ) %>%
  tab_options(
    table.font.size = "small",
    data_row.padding = px(3),
    heading.align = "left"
  ) %>%
  opt_row_striping() %>%
  opt_table_outline()

```

This is a bit of an eye chart but I wanted to show a broad picture of the predictor variables. We can see right away that the variables are not on the same scale, for example chlorides range from 0.01 to 0.61, while total sulfur dioxide ranges from 6 to 440. We may have to standardize predictors. In general red and white wine seem to have similar averages for many variables, at least in the same ballpark. In further exploration and modeling we'll assess what variables and ranges matter most.

### Visualization

```{r}
wine %>%
  pivot_longer(cols = -c(type, quality), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value, fill = type)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~feature, scales = "free") +
  theme_minimal()+
    theme(legend.position = "bottom")+
    labs(title = "Distribution of values across predictor variables", 
         x = "")

```

The density plots show the distribution of each predictor by wine type. I notice a few things right away, such as higher volatile acidity in red wines, and lower sulfur in red wines.

```{r }
wine %>%
  select(-type) %>%
  cor() %>%
  round(2) %>%
  ggcorrplot::ggcorrplot(
    lab = TRUE,
    lab_size = 2.5,       # 🔹 Shrinks the text
    tl.cex = 10,          # 🔹 Shrinks axis labels
    colors = c("red", "white", "blue"),
    outline.color = "gray",
    hc.order = TRUE       # 🔹 Optional: cluster similar variables
  )

```

It doesn't look like there are strong correlations between single variables and quality. Modeling will help us take all variables into consideration.

# Modeling

I will now use linear regression and random forest models to evaluate and compare predictive performance, and identify drivers of wine quality.

### Split the data into test and train

```{r}
set.seed(648)
split <- initial_split(wine, prop = 0.8)
train <- training(split)
test <- testing(split)
```

### Preprocessing

Since we identified the predictors are on different scales, I'm going to standardize them. This dataset started out pretty clean so this is the only preprocessing for now.

```{r}
recipe_obj <- recipe(quality ~ ., data = train) %>% 
    step_normalize(all_numeric_predictors())
```

### Inspect the preprocessed data

I want to inspect the dataset to make sure my preprocessing did what was intended.

```{r}
prep_recipe <- prep(recipe_obj)
baked_train <- bake(prep_recipe, new_data = NULL)
glimpse(baked_train)
```

Looks good for now.

# Specify first model, linear regression

```{r}
model_spec <- linear_reg() %>% 
    set_engine("lm")
```

### Workflow set

```{r}
workflow_obj <- workflow() %>% 
    add_model(model_spec) %>% 
    add_recipe(recipe_obj)
```

### Cross validation

```{r}
set.seed(486)
folds <- vfold_cv(train, v = 5)
```

### Fit the workflow with cross validation

```{r}
fitted_resamples <- workflow_obj %>%
  fit_resamples(
    resamples = folds,
    metrics = metric_set(rmse, rsq),
    control = control_resamples(save_pred = TRUE)
  )

```

### Evaluate performance

```{r}
collect_metrics(fitted_resamples)
```

### Visualize predictions (linear regression)

```{r}
collect_predictions(fitted_resamples) %>%
  ggplot(aes(x = .pred, y = quality)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Predicted vs Actual (CV)", x = "Predicted", y = "Actual") +
  theme_minimal()
```

As I expected, this doesn't seem like a great fit - the RMSE average of 0.73 indicates the model isn't too accurate. The standard error of 0.008 indicates it is performing consistently across the cross validation folds which is good - the model is stable between folds but overall not accurate.

## Final fit and test set predictions

### Fit Final Model

```{r}
final_fit <- workflow_obj %>% 
    fit(data = train)
```

### Test set predictions

```{r}
test_preds <- predict(final_fit, new_data = test) %>% 
    bind_cols(test)
```

### Final performance evaluation

```{r}
test_metrics <- test_preds %>%
  metrics(truth = quality, estimate = .pred)

test_metrics

```

### Predictions visualized

```{r}
test_preds %>%
  ggplot(aes(x = .pred, y = quality)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Predicted vs Actual (Test Set)", x = "Predicted", y = "Actual") +
  theme_minimal()

```

```{r}
test_preds %>%
  mutate(residual = .pred - quality) %>%
  ggplot(aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Predicted", y = "Residuals") +
  theme_minimal()

```

The RMSE was very close for the training and test data, but overall this is not an accurate model. The residual plot shows strange patterns, this model is clearly underfitting.

### Model inspection

```{r}
lm_model <- extract_fit_parsnip(final_fit)$fit
summary(lm_model)
```

Most variables are statistically significant, as is the model itself. I want to point this out and emphasize it, because the "significance" doesn't mean we have a useful model. The model doesn't predict very well.

In the future I might try this with interactions with type - the predictor variable impact might very well depend on the type of wine. Or maybe it's better to fit a model separately for red and white wines.

### Discussion - Linear Regression

Linear regression probably isn't the best choice for the quality rating. We could treat the rating as a categoric variable and do ordinal (multilevel) logistic regression. This is an approach I've taken in some projects. For this project I'll use a random forest model and compare. 

It's worth noting that the quality rating really is a categoric scale, I've been treating it as a continuous number in this project. My reason for doing this, is that this is very often done in industry. I might not agree with it but it's common. We'll take the categoric approach in a separate project.


# Random Forest - next model

For the quality rating, a random forest model will probably perform better than linear regression.

The data is already split into test and train.

(And as a note, I'd use workflowsets if I was doing more of this, to test many models and compare at scale)

I also realized my naming was not the clearest, rather than go back and edit the linear model section, going forward I'm going to denote random forest with _rf.

### Specify random forest model

```{r}
spec_rf <- rand_forest(mode = "regression") %>%
  set_engine("ranger")
```

### Recipe

No preprocessing needed for random forest.

```{r}
rec_rf_wine <- recipe(quality ~ ., data = train)
```

### Workflow

```{r}
wf_rf <- workflow() %>%
  add_model(spec_rf) %>%
  add_recipe(rec_rf_wine)
```

### Cross validation with same folds as lm

```{r}
rf_fit_cv <- fit_resamples(
  wf_rf,
  resamples = folds,
  metrics = metric_set(rmse, rsq),
  control = control_resamples(save_pred = TRUE)
)

rf_results_cv <- collect_metrics(rf_fit_cv)

rf_results_cv
```

I notice the RMSE is lower than with linear regression, the standard error is also lower, and r squared is higher.

### Fit random forest model on training data

```{r}
fit_rf <- fit(wf_rf, data = train)
```

### Predictions on the test set

```{r}
preds_rf <- predict(fit_rf, new_data = test) %>%
  bind_cols(test %>% select(quality))
```

### Calculate metrics

```{r}
results_rf <- preds_rf %>%
  metrics(truth = quality, estimate = .pred)

results_rf
```

Similar performace to the training data.

### Visualize results

```{r}
ggplot(preds_rf, aes(x = .pred, y = quality)) +
  geom_point(alpha = 0.6, color = "#2c7fb8") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Random Forest: Predicted vs Actual Wine Quality",
    x = "Actual Quality",
    y = "Predicted Quality"
  ) +
  theme_minimal()
```

### Discussion - Random Forest

The random forest model (with default hyperparameters) resulted in a more accurate model with lower RMSE, however we can see here that there is still a lot of variation in the results. 

# Future work and conclusions

I didn't do any hyperparameter tuning, we might be able to get better predictive performance with tuning. Regression might not be the right approach, since the response (quality) is a rating, it really is a categoric scale. I might try ordinal logicstic regression instead.

It is possible that the predictors might never give more accurate predictions too. There might be many more variables needed, different variables needed that aren't captured here. For example this dataset considered chemical measurements, but if I was doing consumer research I would want to know demographic and purchase behavior of the tasters. If we could segment the consumers we might find the people who really like (or dislike) certain attributes of wines.

It's also possible that predictions will never be much better, at some point there will be natural variation between product and consumer, and we won't know how they will respond. Sometimes the answer we want doesn't exist! (I am not concluding that yet, this was a simple exploratory project, there is much more to work on)






