---
title: "Chickweight dataset site testing"
author: "Ben Jepson"
format: 
    html:
        toc: true
        code-fold: true
        code-summary: "Show code"
engine: knitr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    echo = FALSE,
    fig.width = 10, 
    fig.height = 7
)
```


```{r}
library(tidyverse)
data(ChickWeight)
theme_set(theme_bw())
```


## Introduction

Using an easy dataset to get used to updating quarto website on github pages
Just added some options and changed the YAML

## Data exploration

```{r}
summary(ChickWeight)
```

Ok I changed the text right here

AND I removed the DT::datatable...I didn't really think it belonged here

```{r}
ggplot(data = ChickWeight, 
       aes(x = Time, y = weight, color = Diet)) +
    
  geom_point(alpha = 0.5) +
    
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = "Chick Growth Over Time by Diet",
       x = "Time (days)",
       y = "Weight (grams)")

```

Changed theme. I'd rather facet by chick maybe? Future edit this is a test.

## Conclusion and next steps

Round 2, see you (soon)